---
title: "Technology Course Day 2: R and RStudio"
author: "Elisha Cohen"
date: "8/25/2019"
header-includes:
    - \usepackage{hyperref}
    - \usepackage{xcolor}
    - \definecolor{purple}{HTML}{431C7C}
    - \hypersetup{colorlinks,breaklinks,linkcolor=purple,urlcolor=purple,anchorcolor=purple,citecolor=black}
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installation

* Install R software from \href{https://cran.r-project.org/}{CRAN}

* Install \href{https://www.rstudio.com/products/rstudio/download/#download}{RStudio}

* Andrew Heiss has on his \href{https://www.andrewheiss.com/blog/2012/04/17/install-r-rstudio-r-commander-windows-osx/}{blog} install instructions for both Windows and OS X

### Set-up 

1. Set your CRAN mirror: preferences > packages

2. Set your theme and font: preferences > appearance

3. Set your pane layout: preferences > Pane Layout

4. In preferences > General, verify "save workspace on exit" set to `NEVER`

### Packages and Libraries

Try installing the `dplyr` package:

```{r packageinstall, eval = FALSE}
install.packages("dplyr")
```

* There is also a drop-down menu under Tools > Install Packages

Load the `dplyr` package from your computer into your RStudio workspace:

```{r packageload, eval=FALSE}
library(dplyr)
```

### Getting Help for a package

Help file:

```{r helpcommand,eval=FALSE}
help(foo)
?foo ## ?? double question marks needed if package isn't loaded into library
```

Vignettes:

```{r vignettes, eval = FALSE}
vignette() # this will show list of all vignettes
vignette(package = "dplyr") # vignettes included for a specific package
vignette("dplyr") # views vignette
```

### Books and other resources

* We will be following R for Data Science (abbreviated \href{https://r4ds.had.co.nz/introduction.html}{R4DS}) which is free online but you can also buy the book if you prefer

* Library has another good resources: \href{http://discovere.emory.edu/primo_library/libweb/action/display.do?tabs=detailsTab&ct=display&fn=search&doc=01EMORY_ALMA51366131950002486&indx=1&recIds=01EMORY_ALMA51366131950002486&recIdxs=0&elementId=0&renderMode=poppedOut&displayMode=full&frbrVersion=&vid=discovere&institution=EMORY&vl(4439115UI0)=any&query=any%2Ccontains%2Chands+on+programming+with+R&libsearch=woodruff&dscnt=0&scp.scps=scope%3A%28repo%29%2Cscope%3A%2801EMORY_ALMA%29%2CEmory_PrimoThirdNode&vl(freeText0)=hands%20on%20programming%20with%20R&dstmp=1566767454149}{Hands-on programming with R} by Garrett Grolemund 
  
### The tidyverse

We will be using a lot of packages in the tidyverse. Install or check for updates:

```{r tidyverse, eval = FALSE}
install.packages("tidyverse")
tidyverse_update()
```

Additional packages used in the R4DS book:

```{r installmmore,eval = FALSE}
install.packages(c("nycflights13", "gapminder", "Lahman"))
```

  