---
title: "R Notes March 25, 2020"
author: "Elisha Cohen"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Matrices in R!

We can start by creating a 3X3 matrix and filling it with the numbers 1 to 9:

```{r c1}
A <- matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE)
A
```

If we use the `class()` function to check the class it should be of the matrix class:

```{r c2}
class(A)
```
This can be important because if matrix operations aren't working it may be that our object is a data.frame or tibble.

Many of the commands we are used to using on other objects continue to work with matrices. We can add a row or add a column:

```{r c3}
## add a row
A <- rbind(A,c(10,11,12))
A
## add a column
A <- cbind(A,c(10,20,30,40))
A
```

To access specific elements, rows or columns we use square brackets always specifying row first then column. Currently our `A` matrix has 4 rows and 4 columns.
```{r c4}
# element in the 2nd row, 3rd column
A[2,3]
```
To return the mth row:
```{r}
# return the entire 3rd row
A[3,]
```
To return the nth column:
```{r}
# return the entire 4th column
A[,4]
```
If we want to multiply the matrix by a scaler:
```{r}
A*2
```
Let's make a 2nd matrix, this time a 4X4 identity matrix:
```{r}
B <- diag(c(1,1,1,1))
```
To multiply matrices use `%*%`
```{r}
A%*%B
```
Note, that the dimensions must match just like with normal matrix operations. So if we have a matrix that is only 3X3 we will get an error that the arguments are non-comformable.
```{r}
C <- matrix(rep(1,8), nrow = 2, ncol = 4)
A%*%C
```
We can transpose a matrix
```{r}
C <- t(C)
```
Now because we have a matrix that is 4X4 and a matrix that is 4X2 our matrix multiplication will work
```{r}
A%*%C
```
To return the inverse of a matrix use solve
```{r}
A <- matrix(1:4,nrow=2,ncol=2)
A
solve(A)
```


