---
title: "Instrumental Variables"
author: "Elisha Cohen"
date: "4/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
path <- file.path("/Users/elisha/Dropbox/current-research/R-Course/POLS509/psets")
```

We'll follow along with the example used in Chapter 3 of Mostly Harmless. You can download the 1980 census from https://economics.mit.edu/faculty/angrist/data1/data/angkru95. Load the data along with `tidyverse()`.

```{r,message=F,warning=F}
library(tidyverse)
library(R.utils) #for file path (you may not need this)
library(AER)
library(texreg) #for nice tables
library(sandwich)
library(estimatr)
library(lfe)
```

```{r,message=F,warning=F,cache=T}
census <- read_delim(filePath(path,"asciiqob.txt"),
                   delim = " ",
                   col_names = c("lwklywge","educ","yob","qob","pob"),
                   n_max = 329509)
# remove extra spaces using stringr package
# and convert to numeric from character
census <- census %>%
  mutate(lwklywge = as.numeric(str_trim(lwklywge, side = "left")),
         educ = as.numeric(str_trim(educ, side = "left")),
         yob = as.numeric(str_trim(yob, side = "left")),
         qob = as.numeric(str_trim(qob, side = "left")),
         pob = as.numeric(str_trim(pob, side = "left")))

# create variable that is year and quarter of birth
census$yqob <- as.numeric(paste0(census$yob,".",census$qob))
# average weekly wage
avg <- census %>%
  group_by(yqob) %>%
  summarize(avg_wkwg = mean(lwklywge),
            avg_educ = mean(educ))
```


Average education by quarter of birth

```{r}
ggplot(data = avg, aes(x = yqob, y = avg_educ)) +
  geom_point() +
  geom_line() +
  ggtitle("Average Education by Quater of Birth (First Stage)") +
  xlab("Year of Birth") +
  ylab("Years of Education") +
  theme_bw()

```


Average weekly wage by quarter of birth

```{r}
ggplot(data = avg, aes(x = yqob, y = avg_wkwg)) +
  geom_point() +
  geom_line() +
  ggtitle("Average Weekly Wage by Quartter of Birth (Reduced Form)") +
  xlab("Year of Birth") +
  ylab("Log Weekly Earnings") +
  theme_bw()

```

### OLS estimation

```{r}
ols <- lm(lwklywge ~ educ, data = census)
ols_robust <- lm_robust(lwklywge ~ educ, data = census, se_type = "HC1",ci=FALSE)
coeftest(ols, vcov=vcovHC, method="HC1")
sqrt(vcovHC(ols, type = "HC1")) #to compare SEs
```

```{r}
summary(ols)
summary(ols_robust)
```

### 2SLS estimation

```{r}
## dummy for QOB == 1
census$q1 <- ifelse(census$qob==1,1,0)
iv <- ivreg(lwklywge ~ educ | q1, data = census)
```

```{r,results='asis'}
htmlreg(list(ols,iv),doctype = F, digits=3,custom.model.names = c("OLS","IV"))
```

### Looking at the Inclusion Restriction

Is our instrument highly correlated with the endogenous regressor? We manually run the first stage and make sure to correct our standard errors for possible heteroscedasticity.

```{r}
iv_stage1 <- lm(educ ~ q1, data = census)
coeftest(iv_stage1, vcov = vcovHC, type = "HC1")
# we could compare to results using the lfe package
test <- felm(lwklywge ~ 1 | 0 |(educ ~ q1)| 0,data=census)
#summary(test$stage1)$coefficients
```

Continue to estimate 2SLS using the predicted values from the first stage.

```{r}
census$educ_pred <- iv_stage1$fitted.values
tsls <- lm(lwklywge ~ educ_pred, data = census)
coeftest(tsls,vcov=sandwich)
coeftest(iv, vcov=sandwich)
```
Compare results between using the `AER` package and doing 2SLS by hand. We should see that the coefficients are the same but the standard errors are different.

```{r,results='asis'}
iv_se_robust <- coeftest(iv, vcov=sandwich)[,"Std. Error"]
tsls_se_robust <- coeftest(tsls, vcov=sandwich)[,"Std. Error"]
htmlreg(list(iv,tsls),
        custom.model.names = c("ivreg","by hand"),
        doctype = F,
        override.se = list(iv_se_robust,tsls_se_robust),
        digits = 3)
```


